---
title: "Recommender System"
author: "Avnish"
date: "30/04/2020"
output: html_document
---
# Introduction
We are building a movie recommender system using open source database *10M version of the MovieLens dataset*.
We have built multiple machine learning models to achieve maximum accuracy.
This document will walk you through all the findings and how we have built different models and rationale behind it.


# Data Partition Test/validation 
```{r, warning=FALSE,message=FALSE}
library(tidyverse)
library(caret)
library(data.table)
library(recosystem)
library(lubridate)

dl <- tempfile()
download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", dl)

ratings <- fread(text = gsub("::", "\t", readLines(unzip(dl, "ml-10M100K/ratings.dat"))),
                 col.names = c("userId", "movieId", "rating", "timestamp"))



movies <- str_split_fixed(readLines(unzip(dl, "ml-10M100K/movies.dat")), "\\::", 3)
colnames(movies) <- c("movieId", "title", "genres")
movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],title = as.character(title),genres = as.character(genres))

movielens <- left_join(ratings, movies, by = "movieId")

set.seed(1, sample.kind="Rounding")
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 0.1, list = FALSE)
edx <- movielens[-test_index,]
temp <- movielens[test_index,]

validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

rm(dl, ratings, movies, test_index, temp, movielens, removed)
```

We will explore the rating distribution

```{r, warning=FALSE,message=FALSE}
edx %>% group_by(rating) %>%
  summarize(count = n()) %>%
  arrange(desc(count))
```


# Variable transformation . 
Exracting release year, rated year and adding 2 varibale releaseyear'and 'ratedafter'.

```{r, warning=FALSE,message=FALSE,cache=TRUE}
edx$timestamp=as.POSIXct(edx$timestamp, origin="1970-01-01")
edx$timestamp=format(as.Date(edx$timestamp, format="%d/%m/%Y"),"%Y")

release = str_sub(edx$title,start = -5,end = -2)
edx=edx%>%mutate(releaseyear=release)
rated=as.numeric(edx$timestamp)-as.numeric(edx$releaseyear)
edx=edx%>%mutate(ratedafter=rated)


```
Number of movies and users

```{r, warning=FALSE,message=FALSE}
edx%>%summarize(users=length(unique(userId)),movies=length(unique(movieId)))

```
Plot Distribution of ratings

```{r, warning=FALSE,message=FALSE}
edx%>%ggplot(aes(rating))+geom_bar()+ggtitle("Rating Distribution")

```
## Plot Distribution of users
We see that there are users more more active than others.

```{r, warning=FALSE,message=FALSE}

edx%>%count(userId)%>%
 ggplot(aes(n))+geom_histogram(bins  = 30,color="red")+
  scale_x_log10()+ggtitle("Distribution of rating providing users")

```
## Distribution of no of times movie has been rated
We see that some movies are watched/rated more than others.


```{r, warning=FALSE,message=FALSE}

edx%>%count(movieId,title)%>%
  ggplot(aes(n))+geom_histogram(bins=30,color="red")+
  scale_x_log10()+ggtitle("Distribution of movie views")

```
# Top 25 viewd movies
```{r, warning=FALSE,message=FALSE}
head(edx%>%count(movieId,title)%>%arrange(desc(n))%>%select(title),25)

```


## How average rating of movies varies for different movies
We see that some movies on average rated higher than others and vice-versa.

```{r, warning=FALSE,message=FALSE}
edx%>%group_by(movieId)%>%summarise(mean=mean(rating))%>%ggplot(aes(mean))+geom_histogram(bins=20,color="red")+ggtitle("Average movie rating across different movies")

```

## How average rating of movies varies for different users
We see that some users on average tend to provide higher rating than other and vice-versa.

```{r, warning=FALSE,message=FALSE}

edx%>%group_by(userId)%>%summarise(mean=mean(rating))%>%
  ggplot(aes(mean))+geom_histogram(bins=20,color="red")+ggtitle("Average movie rating across different users")

```

# How average rating of movies varies for different genres
Average rating also varies across different genre.
```{r, warning=FALSE,message=FALSE}
edx%>%group_by(genres)%>%summarise(mean=mean(rating))%>%
  arrange(desc(mean))%>%
  ggplot(aes(mean))+geom_histogram(bins=20,color="red")+ggtitle("Average movie rating across different genre")

```
Top 25 rated genres

```{r, warning=FALSE,message=FALSE}
head(edx%>%group_by(genres)%>%summarise(mean=mean(rating))%>%
       arrange(desc(mean)),25)
```

## Relation between released year and average rating
Average movie rating also varies across different year, in some years like '1994' movies are rated higher on average

```{r, warning=FALSE,message=FALSE}

edx%>%group_by(releaseyear)%>%summarise(mean=mean(rating))%>%
  ggplot(aes(releaseyear,mean))+geom_point()+geom_smooth()

```

## Relation between rated years after and average rating
How many years after the movies is being rated is also postively related . One of the possible reason will be classis movies.

```{r, warning=FALSE,message=FALSE, cache=TRUE}
edx%>%group_by(ratedafter)%>%summarise(mean=mean(rating))%>%
  ggplot(aes(ratedafter,mean))+geom_point()+geom_smooth()

```

correlation factor 'year release'

```{r, warning=FALSE,message=FALSE}
 
edx%>%group_by(releaseyear)%>%summarise(mean=mean(rating))%>%
  summarize(cor(mean,as.numeric(releaseyear)))

```

correlation factor 'rated year after'

```{r, warning=FALSE,message=FALSE, cache=TRUE}

edx%>%group_by(ratedafter)%>%summarise(mean=mean(rating))%>%
  summarize(cor(mean,ratedafter))

```

## Validation data set prepration

```{r, warning=FALSE,message=FALSE, cache=TRUE}

validation$timestamp=as.POSIXct(validation$timestamp, origin="1970-01-01")
validation$timestamp=format(as.Date(validation$timestamp, format="%d/%m/%Y"),"%Y")

releasev = str_sub(validation$title,start = -5,end = -2)
validation=validation%>%mutate(releaseyear=releasev)
ratedv=as.numeric(validation$timestamp)-as.numeric(validation$releaseyear)
validation=validation%>%mutate(ratedafter=ratedv)

test=validation
validation=validation%>%select(-rating)


```

## RMSE function

```{r, warning=FALSE,message=FALSE, cache=TRUE}
RMSE=function(actual,prediction){
  sqrt(mean((actual-prediction)^2,na.rm = T))
}

```

## Simple Model predicting mean 

```{r, warning=FALSE,message=FALSE, cache=TRUE}
m=mean(edx$rating)
avg_model=RMSE(test$rating,m)

```

Accounting movie effect :

```{r, warning=FALSE,message=FALSE, cache=TRUE}
 
movie=edx%>%group_by(movieId)%>%
  summarise(m_i=mean(rating-m))

```

Accouting user effect :

```{r, warning=FALSE,message=FALSE, cache=TRUE}

user=edx%>%
  left_join(movie,by="movieId")%>%
  group_by(userId)%>%
  summarise(u_i=mean(rating-m-m_i))

```

Accounting release year effect :

```{r, warning=FALSE,message=FALSE, cache=TRUE}

year=edx%>%
  left_join(movie, by="movieId")%>%
  left_join(user,by="userId")%>%
  group_by(releaseyear)%>%
  summarise(y_i=mean(rating-m-m_i-u_i))

```

Accounting genre effect :

```{r, warning=FALSE,message=FALSE}

genre=edx%>%
  left_join(movie, by="movieId")%>%
  left_join(user,by="userId")%>%
  left_join(year,by="releaseyear")%>%
  group_by(genres)%>%
  summarise(g_i=mean(rating-m-m_i-u_i-y_i))

```

Accounting years after rated effect :

```{r, warning=FALSE,message=FALSE}

yearsafter=edx%>%
  left_join(movie, by="movieId")%>%
  left_join(user,by="userId")%>%
  left_join(year,by="releaseyear")%>%
  left_join(genre,by="genres")%>%
  group_by(ratedafter)%>%
  summarise(ya_i=mean(rating-m-m_i-u_i-y_i-g_i))

```

## Model 1: Movie effect

```{r, warning=FALSE,message=FALSE}

pred_by_m=validation%>%
  left_join(movie,by="movieId")%>%
  mutate(prediction=m+m_i)
model_1=RMSE(test$rating,pred_by_m$prediction)

```

## Model2: Movie & User effect

```{r, warning=FALSE,message=FALSE}

pred_by_m_u=validation%>%
  left_join(movie,by="movieId")%>%
  left_join(user, by="userId")%>%
  mutate(prediction=m+m_i+u_i)
model_2=RMSE(test$rating,pred_by_m_u$prediction)

```

## Model3: Movie, User and year effect

```{r, warning=FALSE,message=FALSE}

pred_by_m_u_y=validation%>%
  left_join(movie,by="movieId")%>%
  left_join(user, by="userId")%>%
  left_join(year,by="releaseyear")%>%
  mutate(prediction=m+m_i+u_i+y_i)
model_3=RMSE(test$rating,pred_by_m_u_y$prediction)

```

## Model4: Movie, User, year and genre effect 

```{r, warning=FALSE,message=FALSE}

pred_by_m_u_y_g=validation%>%
  left_join(movie,by="movieId")%>%
  left_join(user, by="userId")%>%
  left_join(year,by="releaseyear")%>%
  left_join(genre,by="genres")%>%
  mutate(prediction=m+m_i+u_i+y_i+g_i)
model_4=RMSE(test$rating,pred_by_m_u_y_g$prediction)


```

## Model5: Movie, User, year, genre and year after rated effect 

```{r, warning=FALSE,message=FALSE}

pred_by_m_u_y_g_ya=validation%>%
  left_join(movie,by="movieId")%>%
  left_join(user, by="userId")%>%
  left_join(year,by="releaseyear")%>%
  left_join(genre,by="genres")%>%
  left_join(yearsafter, by="ratedafter")%>%
  mutate(prediction=m+m_i+u_i+y_i+g_i+ya_i)
model_5=RMSE(test$rating,pred_by_m_u_y_g_ya$prediction)


```

# Regularization

```{r, warning=FALSE,message=FALSE}
lamdas=seq(0,25,1)
error=sapply(lamdas,function(i){
  m=mean(edx$rating)
  
  movie=edx%>%group_by(movieId)%>%
    summarise(m_i=sum(rating-m)/(n()+i))
  
  
  user=edx%>%
    left_join(movie,by="movieId")%>%
    group_by(userId)%>%
    summarise(u_i=sum(rating-m-m_i)/(n()+i))
  
  
  year=edx%>%
    left_join(movie, by="movieId")%>%
    left_join(user,by="userId")%>%
    group_by(releaseyear)%>%
    summarise(y_i=sum(rating-m-m_i-u_i)/(n()+i))
  
  
  genre=edx%>%
    left_join(movie, by="movieId")%>%
    left_join(user,by="userId")%>%
    left_join(year,by="releaseyear")%>%
    group_by(genres)%>%
    summarise(g_i=sum(rating-m-m_i-u_i-y_i)/(n()+i))
  
  
  yearsafter=edx%>%
    left_join(movie, by="movieId")%>%
    left_join(user,by="userId")%>%
    left_join(year,by="releaseyear")%>%
    left_join(genre,by="genres")%>%
    group_by(ratedafter)%>%
    summarise(ya_i=sum(rating-m-m_i-u_i-y_i-g_i)/(n()))
  pred_regularize=validation%>%
    left_join(movie,by="movieId")%>%
    left_join(user, by="userId")%>%
    left_join(year,by="releaseyear")%>%
    left_join(genre,by="genres")%>%
    left_join(yearsafter, by="ratedafter")%>%
    mutate(prediction=m+m_i+u_i+y_i+g_i+ya_i)
  return(RMSE(test$rating,pred_regularize$prediction))
  
})



```

Optimal value of lamda

```{r, warning=FALSE,message=FALSE}
qplot(lamdas,error)
l=lamdas[which.min(error)]
```

## Regulariztion model 6

```{r, warning=FALSE,message=FALSE}

movie=edx%>%group_by(movieId)%>%
  summarise(m_i=sum(rating-m)/(n()+l),n_m=n())
user=edx%>%
  left_join(movie,by="movieId")%>%
  group_by(userId)%>%
  summarise(u_i=sum(rating-m-m_i)/(n()+l),n_u=n())


year=edx%>%
  left_join(movie, by="movieId")%>%
  left_join(user,by="userId")%>%
  group_by(releaseyear)%>%
  summarise(y_i=sum(rating-m-m_i-u_i)/(n()+l),n_y=n())


genre=edx%>%
  left_join(movie, by="movieId")%>%
  left_join(user,by="userId")%>%
  left_join(year,by="releaseyear")%>%
  group_by(genres)%>%
  summarise(g_i=sum(rating-m-m_i-u_i-y_i)/(n()+l),n_g=n())

yearsafter=edx%>%
  left_join(movie, by="movieId")%>%
  left_join(user,by="userId")%>%
  left_join(year,by="releaseyear")%>%
  left_join(genre,by="genres")%>%
  group_by(ratedafter)%>%
  summarise(ya_i=sum(rating-m-m_i-u_i-y_i-g_i)/(n()),n_ya=n())

pred_regularize=validation%>%
  left_join(movie,by="movieId")%>%
  left_join(user, by="userId")%>%
  left_join(year,by="releaseyear")%>%
  left_join(genre,by="genres")%>%
  left_join(yearsafter, by="ratedafter")%>%
  mutate(prediction=m+m_i+u_i+y_i+g_i+ya_i)%>%
  .$prediction
model_6=RMSE(test$rating,pred_regularize)

```

# Model Matrix Factorization using recosystem(LIBMF)
## Data prepration 

```{r, warning=FALSE,message=FALSE}
edx_factor=edx%>%select(movieId,userId,rating)
edx_factor=as.matrix(edx_factor)
write.table(edx_factor,file = "trainset.txt",sep = " ",row.names = FALSE,col.names = FALSE)
trainset=data_file("trainset.txt")


```

Recosystem object creation

```{r, warning=FALSE,message=FALSE}
r=Reco()
```

## Training

```{r, warning=FALSE,message=FALSE}
r$train(trainset, opts = list(dim = 30,nthread = 1,niter=20))
stored_prediction = tempfile()
```

## Validation

```{r, warning=FALSE,message=FALSE}
validation_factor=validation%>%select(movieId,userId)
validation_factor=as.matrix(validation_factor)
write.table(validation_factor,file = "testset.txt",sep = " ",row.names = FALSE,col.names = FALSE)
testset=data_file("testset.txt")
r$predict(testset,out_file(stored_prediction))
prediction=scan(stored_prediction)
```

# RMSE

```{r, warning=FALSE,message=FALSE}
recosystem_model=RMSE(test$rating,prediction)
```

# All model Summary

```{r, warning=FALSE,message=FALSE}
recosystem_model=RMSE(test$rating,prediction)
all_model=data.frame(modelname=c("Simple Model","movie","movie_user","movie_user_year","movie_user_year_genre","movie_user_year_genre_ratedafter","Regularization","Recosystem"),modelperf=c(avg_model,model_1,model_2,model_3,model_4,model_5,model_6,recosystem_model))
all_model
```
